---
title: "Shootings in New York"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    theme: united
---

```{r setup, message=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("readr")
library("dplyr")
library("tidyr")
library("lubridate")
library("tidyverse")
library("ggthemes")
library("knitr")
```

### Introduction

In this project I will primarily be analyzing a dataset of shootings in the New York boroughs in the year 2021.  I have also included some extra supplemental data for the population of the New York boroughs that same year.  I am interested in looking into the number of shootings grouped by race and age group of the perpetrator.  For the models I'd like to analyze the number of shootings and try to determine why they have increased or decreased at times. 

I have shown my data analysis process in the following steps: Gathering, Transforming, Cleaning, Modeling & Analysis, and Conclusion.  

---

### Gathering

First, read in the data:
```{r readData, message=FALSE}
url = "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD"
shootings = read.csv(url)
```

---

### Transforming
Transform the data removing unnecessary columns & converting date to the proper type

```{r transform}

tidyShootings = shootings %>% 
  select(c(OCCUR_DATE, 
           BORO,
           PERP_AGE_GROUP, 
           PERP_SEX,
           PERP_RACE,
           VIC_AGE_GROUP,
           VIC_SEX,
           VIC_RACE)) %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE))
```

---

### Cleaning

The below command will show the unique values in every column (except date).  Note the undesired values like "", "UNKNOWN", and incorrect ages. 
```{r uniqueValues}
# The function "sapply" (similar to "lapply" but returns a vector instead of a list) applies the command in () to each column
# By using " - " in front of the collection we are specifying every column except OCCUR_DATE
tidyShootings %>% 
  select(-c(OCCUR_DATE)) %>%
  sapply(unique)
```

Rows to remove:

  1. PERP_AGE_GROUP: "", 1020, 940, 224, UNKNOWN
  2. PERP_SEX: ""
  3. PERP_RACE: "", "UNKNOWN"
  4. VIC_AGE_GROUP: "UNKNOWN" 
  5. VIC_RACE: "UNKNOWN"
```{r filter}
# The logical operator ! inverts the boolean result of the conditional expression present in the ()
# "PERP_AGE_GROUP %in% c(...)" allows us to compare the entries of PERP_AGE_GROUP with all the values present in the vector at once instead of listing them on multiple lines
tidyShootings <- tidyShootings %>%
  filter(!(PERP_AGE_GROUP %in% c("", "1020", "940", "224", "UNKNOWN"))) %>%
  filter(PERP_SEX != "") %>%
  filter(PERP_RACE %in% c("BLACK", "ASIAN / PACIFIC ISLANDER", "BLACK HISPANIC", "WHITE HISPANIC", "WHITE", "AMERICAN INDIAN/ALASKAN NATIVE")) %>%
  filter(VIC_AGE_GROUP != "UNKNOWN") %>%
  filter(VIC_RACE != "UNKNOWN")

```

---

### Modeling & Analyzing
The following chart shows the number of shootings in New York broken down by the race of the perpetrator, and it's an interesting graph for sure, but the large numbers in the 18-24 & 25-44 categories make it difficult to see the 45-64 & 65+ comparisons.  Let's try looking at it another way.
```{r analysisPERPsCount, fig.width = 10}
# theme_wsj = Theme used by the Wall Street Journal :)
ggplot(data = tidyShootings) + 
  geom_bar(mapping = aes(x = PERP_AGE_GROUP, fill = PERP_RACE), position = "dodge") +
  labs(title = "Count of Perpetrators", subtitle = "Number of perpetrators by age group & race", fill = "Race", caption = "Source: Data.gov") + 
  theme_wsj()

```

The next chart uses proportions, giving us a better breakdown of the perpetrator race, but this example can also be misleading.  When viewed independently, the final age group, 65+, seems to show a growth in white perpetrators, but returning to the previous graph you can see the number of white perpetrators hasn't changed significantly.  You need to view both the following and previous graphs together to understand the data so that you are not mislead by one independently.

```{r analysisPERPsProp, warning = FALSE, fig.width = 10}
# theme_wsj = Theme used by the Wall Street Journal :)
ggplot(data = tidyShootings) + 
  geom_bar(mapping = aes(x = PERP_AGE_GROUP, fill = PERP_RACE), position = "fill") +
  labs(title = "Proportions of Perpetrators", subtitle = "Proportions broken down by age group & race", fill = "Race", caption = "Source: Data.gov") + 
  theme_wsj()

```

The data from the previous visualizations shows a disproportionate distribution of shooters between the races present in the datatset.  This could be caused if there isn't a proportional distribution of races in the New York boroughs. To verify this hypothesis I have retrieved the population estimates of the boroughs from 2021 (see data sources for reference).  

Note: I was unable to find the exact race division present in our original dataset, but it is close enough to understand the general idea.  

The data confirms our theory that the races are not proportional even in the entire city let alone throughout the boroughs.  However, this brings up more questions.  The population proportion shows White people make up almost half of the New York borough populations (46.773%), but in our dataset they represent less than a quarter (17.667%).  On the other hand, the results for Black people throughout the New York boroughs is then the opposite, they make up 30.294% of the city's population, but 81.307% of our dataset.  

This could be the result of underlying biases during data collection (e.g., reporting bias, confirmation bias, etc.), but without further samples it is impossible to know for sure.


```{r analysisProportions}
# tibble is used to create a new dataset
# I have retrieved these numbers from an external source - see data sources for reference
population_2021 <- tibble(
  Race = c(
    "AMERICAN INDIAN/ALASKAN NATIVE", "ASIAN / PACIFIC ISLANDER", "BLACK", "WHITE"),
  Population = c(86218, 1385144, 1943645, 3000945),
  Population_Proportion = paste(round((Population/sum(Population)) * 100, 3), '%')
) 

# The kable function produces a nicer table output than the base R tables
kable(population_2021)

# Doing some math in the mutuate lines to calculate the proportions, then clean it up and add '%'
grouped <- tidyShootings %>%
  group_by(Perpetrator_Race = PERP_RACE) %>%
  summarize(Occurences = n()) %>%
  mutate(Sample_Proportion = round((Occurences / sum(Occurences)) * 100, 3)) %>%
  mutate(Sample_Proportion = paste(Sample_Proportion, '%'))

# The kable function produces a nicer table output than the base R tables
kable(grouped)
```

This final visualization is a good representative by itself since the count of incidents in any single borough is not drastically higher than the rest (like the first graph we saw).  With this we can get a good idea of the number of shootings in each borough from one chart.
```{r analysisBOROs, warning = FALSE, fig.width = 10}
# theme_wsj = Theme used by the Wall Street Journal :)
ggplot(data = tidyShootings) +
  geom_bar(mapping = aes(x = BORO, fill = BORO)) +
  labs(title = "Danger in New York", subtitle = "Shootings broken down by borough", caption = "Source: Data.gov") +
  guides(fill = FALSE) +
  theme_wsj()

```

The next model shows the shootings in the New York boroughs over the years.  It's interesting to note the uptick in 2019 presumably due to Covid 19.  

```{r modelCovid, fig.width = 10, message = FALSE}
by_date <- tidyShootings %>%
  group_by(Year = year(OCCUR_DATE), BORO) %>%
  summarize(count = n()) 

ggplot(data = by_date) + 
    geom_point(aes(x = Year, y = count, color = BORO)) + 
    geom_line(aes(x = Year, y = count, color = BORO)) + 
    annotate(geom = "text", x = 2018, y = 225, label = "Covid19", color = "red") + 
    annotate(geom = "rect", xmin = 2018.5, xmax = 2019.5, ymin = 0, ymax = 200, fill = "red", alpha = .2) +
    labs(title = "Shootings in New York", subtitle = "Note the sharp rise after 2019", caption = "Source: Data.gov", color = "Borough") + 
    theme_wsj()
    
```

Finally, this last model smooths out the lines from the previous graph so we can try to see any patterns.  This could be used to loosely predict future shooting numbers or to estimate shootings in the past for missing data. 
```{r modelError, fig.width = 10, message = FALSE}
ggplot(data = by_date) + 
    geom_smooth(aes(x = Year, y = count, color = BORO)) + 
    labs(title = "Shootings in New York", subtitle = "Noticing the patterns", caption = "Source: Data.gov", color = "Borough") + 
    theme_wsj()
```



---

### Conclusion

With any data analysis, we must remember that datasets are not perfect representations of reality, and there are varying kinds of biases that could be present in every stage of the analysis process.  As an example, if this data was collected via police or media reports, it could have a reporting bias which entails specific kinds of shootings receiving more police / media coverage.  There is also confirmation bias, meaning if the person collecting the data had a bias regarding specific groups of people, they might pay more attention to areas those groups of people reside in over other areas. 

I have tried to remove any occurrences of a personal bias such as an anchoring bias where the analyst will rely on the first impressions of their analyses.  With the population breakdown, I was able to show that there might be a bias present in the data, but without any further datasets, we can't be sure.  

Thank you for your time reviewing my project,

-- E

---

### Data Sources

  1. The first source I used was the Shootings in New York in 2021.  This gives us data of the perpetrators, victims, fatality of the shooting, location, and date:  https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv?accessType=DOWNLOAD
  2. I also included some supplemental data describing the population of New York boroughs in 2021 broken down by race so I could compare it with the proportion of shooters by race: http://www.citypopulation.de/en/usa/newyorkcity/

---

#### Session Info
```{r sessionInfo, echo=FALSE}
sessionInfo()
```






















